The host galaxy of the short GRB~111117A at $z = 2.211$\thanks{Based on observations collected at the European Southern Observatory, Paranal, Chile, Program ID: 088.A-0051 and 091.D-0904.}

### 2017-06-13 GRB111117A.tex
Address Valerios comments
Closes #2

The host lies in the brighter end of the expected host brightness distribution
at $z = 2.211$, and is actively forming stars. Using the host as a benchmark for
redshift determination, we find that less than 64 per cent of all sGRB redshifts
should be missed due to host faintness at $z\sim2$. The high redshift of
GRB~111117A is evidence against a lognormal delay-time model for sGRBs through
the predicted redshift distribution of sGRBs, which is very sensitive to
high-$z$ sGRBs.

@@

The host galaxies of sGRBs are diverse. They are more massive and less actively
star-forming on average than lGRB hosts \citep{Fong2013b}, while in some cases,
no host galaxy can be identified above detection threshold of deep follow-up
observations \citep{Berger2010a, Tunnicliffe2014}. Together with their position
within their hosts \citep{Fong2013a}, this suggests a progenitor system that can
be very long-lived, and is associated with stellar mass rather than
star-formation rate (SFR) \citep{Berger2014}. The median redshift for sGRBs is
$z\sim0.5$ \citep{Berger2014}, and because most of these measurements come from
the associated hosts, it is arguably biased towards lower redshifts.

So far, the electromagnetic signals from sGRBs are the only means to accurately
localize NS mergers and hold the promise for detection of an associated
gravitational wave signal \citep{Ghirlanda2016}. From the burst and following
afterglow, modeling the properties and mapping the environments additionally
allows insights into the phenomenon itself.

The total lifetimes of NS binaries depends on their orbit, mass, spin, initial
separations and subsequent inspiral times. The delay time from formation to
explosion impacts the timing and distribution of the enrichment of the ISM with
heavy $r$-process elements \citep{VandeVoort2015, Wallner2015,  Ji2016}. Some
limits can be calculated based on models of star-formation histories of, and the
spatial distribution of sGRBs in, their host galaxies \citep[][]{Berger2014}.
The most distant cosmological bursts, however, offer direct, hard limits on the
coalescence time scales.

@@

\begin{figure}
	\centering
	\includegraphics[width=9cm]{figures/GRB111117A_spec_obs.pdf}
	\caption{
	Imaging of the field of GRB~111117A with the X-shooter slit overlaid. The slit
	position represents 4 epochs of spectroscopic observations taken at similar
	position angles. The image is the FORS2 $R$-band image for which the photometry
	is shown in Fig.~\ref{fig:SED}. The blue asterisk indicates the GRB position as
	derived from the \emph{Chandra} observations in \citet{Sakamoto2013}. 
%	The insert shows the 2D-image of the \oiii$\lambda$5007 emission line. The
%	location of a bright skyline is marked by the blue box. The location of the
%	emission line is indicated with the red ellipse. Because the host is observed
%	in nodding-mode, negative images of the emission line appear on both sides in
%	the spatial direction.
	}
	\label{fig:spec_setup}
\end{figure}

@@

We determine a redshift of $z = 2.211$ from the simultaneous detection of
emission lines belonging to \lya, \oii, \hb, \oiii$\lambda$5007, and \ha. 
\hb~is detected at low significance ($\sim 3 \sigma$). We show \oiii$\lambda$5007
in the insert in Fig.~\ref{fig:line}.  \ha~is only visible in the first
epoch due to the $K$-band blocking filter used for the remaining observations.
The nebular lines exhibit a spatial extent of $\sim$ 1\farc5 and show
significant velocity structure along the slit. A drop in the continuum to the
blue of the \lya~line further supports the inferred redshift.

The total extent of the lines in velocity space is $\sim$ 450 km/s. The line
 profiles shows an asymmetric "double-horned" profile, indicating that we are
 seeing a galaxy with a large degree of coherent rotational motion relative to
 the line-of-sight. If we assume that we are viewing a spiral galaxy exactly
 edge-on, this is a measure of the rotational velocity of the gas. If we assume
 that the spectral resolution and the turbulent width of the lines are
 negligible compared to the rotational velocity, we can based on the projected
 size of the source and the width of the lines, put a constraint on the
 dynamical mass of the galaxy. Based on the physical scale from extent and the
 velocity with, we infer M$_\text{dyn} \gtrsim 10^{10.8}$ M$_\odot$. Because we
 are viewing the host at an angle, this value is an upper limit, which should
 be divided by $\sin(i)$, where $i$ is the inclination relative to edge-on.

Using the luminosity of \ha, we can infer the star-formation rate (SFR) of a
galaxy \citep{Kennicutt1998}. At the redshift of the GRB host, \ha~is observed
at 21\,000~\AA~where the night sky is very bright. In addition, several bright
sky-lines are superposed on the line, making an accurate estimate of the
\ha-flux difficult. We obtain a limit on the SFR by integrating the part of
\ha~free of contamination and correcting for the missing flux using the line
shape. After converting the \citet{Kennicutt1998} relation to a
\citet{Chabrier2003} initial mass function using \citet{Madau2014}, we derive a
limit of $SFR > 7 M_\odot$~yr$^{-1}$. From the SED-fit (Sect.~\ref{SED}), and
the detection of \lya, the host is constrained to contain very little or no
dust, although the presence of Lya does not exclude dust. Therefore we do not
apply a dust-correction to the measured \ha~flux here. Secondly, based on the
width of \oiii, an aperture covering the line is integrated over, where
synthetic sky lines \citep{Noll2012, Jones2013} have been masked and
interpolated over. From the integrated \ha-line, we estimate SFR = $18 \pm 3$
M$_\odot$ yr$^{-1}$.

@@

The multi-color SED is fit by \citet{Bruzual2003} single stellar population
models based on a \citet{Chabrier2003} with initial mass function in
\emph{LePhare} \citep{Ilbert2006}, where the redshift is fixed to the
spectroscopic value of $z=2.211$. The best fit model is an unreddened galaxy
template, and returns physical parameters of absolute magnitude
($M_B=-22.0\pm0.1$\,mag), stellar mass ($\log(M_{\star}/M_\odot) = 9.9\pm0.2$),
stellar population age ($\tau = 0.5_{-0.3}^{+0.5}$ Gyr) and star-formation rate
($SFR_{\mathrm{SED}}=11_{-4}^{+9} M_\odot\,\mathrm{yr}^{-1}$). We show the SED
fit in Fig.~\ref{fig:SED}.

@@

As already pointed out \citep{Margutti2012, Sakamoto2013}, GRB~111117A is
securely classified as belonging to the short class of GRBs. Because the
observed classification indicators, $T_{90}$ and hardness ratio, do not depend
strongly on redshift \citep{Littlejohns2013a}, the updated redshift does not
change this designation.
The intrinsic luminosity is higher as demonstrated in the X-ray light curve
(Fig. \ref{fig:sxray_lightcurve}), but it is still is sub-luminous compared to
the majority of long GRBs.
\citet{Bromberg2013} investigated the degree to which the long and short
population distributions overlap and quantified the certainty in the class
membership. According to \citet{Bromberg2013}, GRB~111117A has $96_{-5}^{+3}$
percent probability of being a sGRB. Compared to the other two sGRB candidates
at high redshift, GRB~060121 \citep{DeUgartePostigo2006, Levan2006} at $1.7
\lesssim z \lesssim 4.5$ ($17_{-15}^{+14}$ per cent) and GRB~090426
\citep{Antonelli2009, Levesque2010, Thone2011} at $z = 2.609$ ($10_{-10}^{+15}$
per cent), the certainty in class membership for GRB~111117A is much higher.
This means that GRB~111117A happened when the universe was younger by 1.8 Gyr
compared to any other non-collapsar GRB ever detected. This number is 3.2 Gyr
for the next-highest spectroscopic redshift. This has implications for the
$r$-process element enrichment timescale of the universe.

@@

We show the recalculated $N_\mathrm{H}$ in Fig.~\ref{fig:NH_z} where we compare
with the distributions of complete samples of both long and short GRBs. The lGRB
sample is from \citet{Arcodia2016} and the sGRB sample is from
\citet{DAvanzo2014a}. 17 out of the 99 long bursts do not have measured redshifts, and
likewise for 5 out of 16 in the short sample. Bursts without measured redshifts have been
excluded for both groups. GRB~111117A occupies a unique position in
Fig.~\ref{fig:NH_z} with the highest $N_H$ of all short bursts. Additionally the
upper limit at $z = 2.609$ belongs to GRB~090426 and does not belong in the
short sample. The short sample, excluding GRB~111117A, is located at low
redshifts ($z < 1$) and is found to populate a similar column density
environment to lGRBs at similar redshifts \citep{DAvanzo2014a}. 
The inferred hydrogen column density for GRB~111117A seems to follow the trend
with increasing $N_H$ as a function of redshift as found for the lGRB afterglows
\citep{Arcodia2016}. This is similar to what is found by \citet{Kopac2012,
	Margutti2013}, that $N_H$ seems to be comparable for long and short GRBs when
compared at similar redshifts.

>>>>

The host lies in the brighter end of the expected sGRB host brightness distribution
at $z = 2.211$, and is actively forming stars. Using the host as a benchmark for
redshift determination, we find that less than 64 per cent of all sGRB redshifts
should be missed due to host faintness for hosts at $z\sim2$. The high redshift of
GRB~111117A is evidence against a lognormal delay-time model for sGRBs through
the predicted redshift distribution of sGRBs, which is very sensitive to
high-$z$ sGRBs.

@@

The \textit{Swift} satellite \citep{Gehrels2004} greatly improved the
understanding of sGRB progenitors thanks to its quick localization capability.
The bulk of these localizations have associated galaxies at relatively low
redshifts with a median redshift of $z\sim0.5$ \citep{Berger2014}, and because
most of these measurements come from the associated hosts, it is arguably biased
towards lower redshifts.

The host galaxies of sGRBs are diverse. They are more massive and less actively
star-forming on average than lGRB hosts \citep{Fong2013b}, while in some cases,
no host galaxy can be identified above detection threshold of deep follow-up
observations \citep{Berger2010a, Tunnicliffe2014}. Together with their position
within their hosts \citep{Fong2013a}, this suggests a progenitor system that can
be very long-lived, and is associated with host stellar mass rather than
star-formation rate (SFR).

The electromagnetic signals from sGRBs are a promising channel to accurately
localize NS mergers, which holds the promise for a detection of an associated
gravitational wave (GW) signal \citep{Ghirlanda2016}. The simultaneous detection
of the sGRB and the GW will provide new a promising ways to constrain the binary
inclination angle \citep{Arun2014} and measure cosmological distances
\citep{Nissanke2010}.

The total lifetimes of NS binaries depends on their orbit, mass, spin, initial
separations and subsequent inspiral times. The delay time from formation to
explosion impacts the timing and distribution of the enrichment of the ISM with
heavy $r$-process elements \citep{VandeVoort2015, Wallner2015,  Ji2016}. Some
limits can be calculated using host galaxy star-formation history models and
spatial distribution of sGRBs within their hosts \citep[][]{Berger2014}. The
most distant cosmological bursts, however, offer direct, hard limits on the
coalescence time scales.

@@

\begin{figure}
	\centering
	\includegraphics[width=9cm]{figures/GRB111117A_spec_obs.pdf}
	\caption{
	FORS2 R-band imaging of the field of GRB~111117A with the X-shooter slit overlaid. The slit
	position represents 4 epochs of spectroscopic observations taken at similar
	position angles. The corresponding photometry
	is shown in Fig.~\ref{fig:SED}. The blue asterisk indicates the GRB position as
	derived from the \emph{Chandra} observations in \citet{Sakamoto2013}. 
%	The insert shows the 2D-image of the \oiii$\lambda$5007 emission line. The
%	location of a bright skyline is marked by the blue box. The location of the
%	emission line is indicated with the red ellipse. Because the host is observed
%	in nodding-mode, negative images of the emission line appear on both sides in
%	the spatial direction.
	}
	\label{fig:spec_setup}
\end{figure}

@@

We determine a redshift of $z = 2.211$ from the simultaneous detection of
emission lines belonging to \lya, \oii$\lambda$3727, \hb, \oiii$\lambda$5007,
and \ha. \hb~is detected at low significance ($\sim 3 \sigma$). We show
\oiii$\lambda$5007 in the insert in Fig.~\ref{fig:line}.  \ha~is only visible in
the first epoch due to the $K$-band blocking filter used for the remaining
observations. The nebular lines exhibit a spatial extent of $\sim$ 1\farc5 and
show significant velocity structure along the slit. A drop in the continuum
bluewards of the \lya~line further supports the inferred redshift.

%The total extent of the lines in velocity space is $\sim$ 450 km/s. The line
% profiles shows an asymmetric "double-horned" profile, indicating that we are
% seeing a galaxy with a large degree of coherent rotational motion relative to
% the line-of-sight. If we assume that we are viewing a spiral galaxy exactly
% edge-on, this is a measure of the rotational velocity of the gas. If we assume
% that the spectral resolution and the turbulent width of the lines are
% negligible compared to the rotational velocity, we can based on the projected
% size of the source and the width of the lines, put a constraint on the
% dynamical mass of the galaxy. Based on the physical scale from extent and the
% velocity with, we infer M$_\text{dyn} \gtrsim 10^{10.8}$ M$_\odot$. Because we
% are viewing the host at an angle, this value is an upper limit, which should
% be divided by $\sin(i)$, where $i$ is the inclination relative to edge-on.

Using the luminosity of \ha, we can infer the star-formation rate (SFR) of a
galaxy \citep{Kennicutt1998}. At the redshift of the GRB host, \ha~is observed
at 21\,000~\AA~where the night sky is very bright. In addition, several bright
sky-lines are superposed on the line, making an accurate estimate of the
\ha-flux difficult. We obtain a limit on the SFR by numerically integrating the part of
\ha~free of contamination and correcting for the missing flux using the line
profile. After converting the \citet{Kennicutt1998} relation to a
\citet{Chabrier2003} initial mass function using \citet{Madau2014}, we derive a
limit of $SFR > 7 M_\odot$~yr$^{-1}$. Additionally, based on the
width of \oiii, an aperture covering the line is integrated over, where
synthetic sky lines \citep{Noll2012, Jones2013} have been masked and
interpolated over. From the integrated \ha-line, we estimate SFR = $18 \pm 3$
M$_\odot$ yr$^{-1}$. From the SED-fit (Sect.~\ref{SED}), and
the detection of \lya, the host is constrained to contain very little or no
dust, although the presence of Lya does not exclude dust. Therefore we do not
apply a dust-correction to the measured \ha~flux here.

@@

The multi-color SED is fit using the \citet{Bruzual2003} single stellar population
models based on a \citet{Chabrier2003} with initial mass function in
\emph{LePhare} \citep{Ilbert2006}, where the redshift is fixed to the
spectroscopic value of $z=2.211$. The best fit model is an unreddened galaxy
template, and returns physical parameters of absolute magnitude
($M_B=-22.0\pm0.1$\,mag), stellar mass ($\log(M_{\star}/M_\odot) = 9.9\pm0.2$),
stellar population age ($\tau = 0.5_{-0.3}^{+0.5}$ Gyr) and star-formation rate
($SFR_{\mathrm{SED}}=11_{-4}^{+9} M_\odot\,\mathrm{yr}^{-1}$). We show the SED
fit in Fig.~\ref{fig:SED}.

@@

As pointed out by \citet{Margutti2012} and \citet{Sakamoto2013}, GRB~111117A is
securely classified as a sGRBs. Because the observed classification indicators,
$T_{90}$ and hardness ratio, do not depend strongly on redshift
\citep{Littlejohns2013a}, the updated redshift does not change this designation.
The intrinsic luminosity is higher as demonstrated in the X-ray light curve
(Fig. \ref{fig:sxray_lightcurve}), but it is still is sub-luminous compared to
the majority of long GRBs. The intrinsic spectral lag shortens, further pushing
the classification in the short regime.

\citet{Bromberg2013} investigated the degree to which the long and short
population distributions overlap and quantified the certainty in the class
membership. According to \citet{Bromberg2013}, GRB~111117A has $96_{-5}^{+3}$
percent probability of being a sGRB. Compared to the other two sGRB candidates
at high redshift, GRB~060121 \citep{DeUgartePostigo2006, Levan2006} at $1.7
\lesssim z \lesssim 4.5$ ($17_{-15}^{+14}$ per cent) and GRB~090426
\citep{Antonelli2009, Levesque2010, Thone2011} at $z = 2.609$ ($10_{-10}^{+15}$
per cent), the certainty in class membership for GRB~111117A is much higher.
This means that GRB~111117A happened when the universe was younger by 1.8 Gyr
compared to any other non-collapsar GRB ever detected. This number is 3.2 Gyr
for the next-highest spectroscopic redshift. This has implications for the
$r$-process element enrichment timescale of the universe.

@@

We show the recalculated $N_\mathrm{H}$ in Fig.~\ref{fig:NH_z} where we compare
with the distributions of complete samples of both long and short GRBs. The lGRB
sample is from \citet{Arcodia2016} and the sGRB sample is from
\citet{DAvanzo2014a}. Both comparison samples consider the largest temporal
interval of constant hardness ratio to prevent spectral changes that can affect
the X-ray derived column density. The 17 (5) of the 99 (16) long (short) GRBs
which do not have redshift have been excluded from our analysis.

GRB~111117A occupies a unique position in Fig.~\ref{fig:NH_z} with the highest
$N_H$ of all short bursts. Additionally the upper limit at $z = 2.609$ belongs
to GRB~090426 and does not belong in the short sample. The short sample,
excluding GRB~111117A, is located at low redshifts ($z < 1$) and is found to
populate a column density environment similar to that of lGRBs at comparable
redshifts \citep{DAvanzo2014a}. The inferred hydrogen column density for
GRB~111117A seems to follow the trend with increasing $N_H$ as a function of
redshift as found for the lGRB afterglows \citep{Arcodia2016}. This is similar
to what is found by \citet{Kopac2012, Margutti2013}, that $N_H$ seems to be
comparable for long and short GRBs when compared at similar redshifts.
